---
title: "Agricultural Ecology Obsservation Disparity Analyzer"
output:
  html_document:
    highlight: pygments
    theme: cerulean
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---
# Instructions

Given a data set of observations, how many were collected on agricultural lands vs non-agricultural lands?

## Create a GBIF Data set

What taxa, time frame, and geography are we interested in?

1.  Visit - <https://www.gbif.org/occurrence/search>

2.  Use the interface to filter down a data set and then press download
    to create a unique DOI

3.  Compiled data sets and DOI codes are created here:
    <https://www.gbif.org/user/download>

4.  Download your data set as a CSV.

## Download Land Use Data sets

-   USGS NLCD 2021 - NLCD Land Cover (CONUS) All Years -
    [https://www.mrlc.gov/data](https://www.google.com/url?q=https%3A%2F%2Fwww.mrlc.gov%2Fdata)

-   USDA NASS CroplandCROS - National Download -
    [https://www.nass.usda.gov/Research_and_Science/Cropland/Release/index.php](https://www.google.com/url?q=https%3A%2F%2Fwww.nass.usda.gov%2FResearch_and_Science%2FCropland%2FRelease%2Findex.php)

#Setup
## Set Variables

```{r}
#GBIF SUBSET
gbif_dataset_path = 'Datasets/GBIF/GBIF_SLF/slf.csv'

#USGS NLCD 2021
nlcd_dataset_path = 'Datasets/Land Use/NLCD2021/nlcd_2021_land_cover_l48_20230630/nlcd_2021_land_cover_l48_20230630.img'

#USDA NASS CroplandCROS
nass_dataset_path = 'Datasets/Land Use/NASS_USA_2023/2023_30m_cdls/2023_30m_cdls.tif'
```

## Install Required Packages

```{r}
# Install necessary packages if not already installed
if(!require(terra)) install.packages("terra")
if(!require(dplyr)) install.packages("dplyr")
if(!require(readr)) install.packages("readr")
if(!require(ggplot2)) install.packages("ggplot2")

# Load the required libraries
library(terra)
library(dplyr)
library(readr)
library(ggplot2)
```

## Load and Clean Datasets

```{r}
# Load USGS NLCD raster dataset
nlcd_raster <- rast(nlcd_dataset_path)

# Load USDA NASS CroplandCROS raster dataset
nass_raster <- rast(nass_dataset_path)

# Load GBIF data set as a data frame
gbif_data <- read_csv(gbif_dataset_path)

# Check if both raster datasets have the same CRS
if (crs(nlcd_raster) == crs(nass_raster)) {
  cat("The CRS of both rasters are the same.\n")
} else {
  cat("The CRS of the rasters are different.\n")
  cat("NLCD CRS:", crs(nlcd_raster), "\n")
  cat("NASS CRS:", crs(nass_raster), "\n")
}

# Plot the rasters to visualize
plot(nlcd_raster, main = "USGS NLCD 2021")
plot(nass_raster, main = "USDA NASS CroplandCROS")
```

## Clean and Prepare GBIF Dataset
```{r}
# Load GBIF dataset
gbif_data <- read_csv(gbif_dataset_path)

# Filter out rows with missing or invalid longitude/latitude
gbif_data <- gbif_data %>%
  filter(!is.na(decimalLongitude) & !is.na(decimalLatitude)) %>%
  filter(decimalLongitude >= -180 & decimalLongitude <= 180,
         decimalLatitude >= -90 & decimalLatitude <= 90)

# Convert GBIF data points to SpatVector using WGS 84 CRS (assuming GBIF uses WGS 84)
coordinates <- vect(gbif_data, geom = c("decimalLongitude", "decimalLatitude"), crs = "EPSG:4326")
# Project GBIF points to match the CRS of the NLCD raster
coordinates_nlcd <- project(coordinates, crs(nlcd_raster))
# Project GBIF points to match the CRS of the NASS raster
coordinates_nass <- project(coordinates, crs(nass_raster))
```

# Analysis

## Analyze Observations Against Land Use
```{r}
# Extract values from the NLCD raster at the reprojected GBIF points
gbif_data$nlcd_value <- extract(nlcd_raster, coordinates_nlcd)[, 2]

# Extract values from the NASS raster at the reprojected GBIF points
gbif_data$nass_value <- extract(nass_raster, coordinates_nass)[, 2]
```


```{r}
# Show all possible values for each column
# List all unique values in the nlcd_value column
unique_nlcd_values <- unique(gbif_data$nlcd_value)
print(unique_nlcd_values)

# List all unique values in the nass_value column
unique_nass_values <- unique(gbif_data$nass_value)
print(unique_nass_values)

# Define land use codes for NLCD and NASS datasets
# Manually define numeric codes for NLCD labels
nlcd_land_use_codes <- c(
  "Deciduous Forest"            = 0,
  "Evergreen Forest"            = 0,
  "Mixed Forest"                = 0,
  "Developed, Low Intensity"    = 0,
  "Developed, Medium Intensity" = 0,
  "Developed, High Intensity"   = 0,
  "Developed, Open Space"       = 0,
  "Open Water"                  = 0,
  "Woody Wetlands"              = 0,
  "Emergent Herbaceous Wetlands" = 0,
  "Cultivated Crops"            = 1,
  "Hay/Pasture"                 = 1,
  "Shrub/Scrub"                 = 0,
  "Barren Land"                 = 0,
  "Herbaceous"                  = 0,
  "Unclassified"                = 0  # Assuming 0 for unclassified
)

# Manually define numeric codes for NASS labels
# Expanded NASS Land Use Codes
nass_land_use_codes <- c(
  "Deciduous Forest"          = 0,   # Non-agricultural
  "Woody Wetlands"            = 0,   # Non-agricultural
  "Developed/High Intensity"  = 0,   # Non-agricultural
  "Developed/Med Intensity"   = 0,   # Non-agricultural
  "Developed/Open Space"      = 0,   # Non-agricultural
  "Corn"                      = 1,   # Agricultural
  "Developed/Low Intensity"   = 0,   # Non-agricultural
  "Mixed Forest"              = 0,   # Non-agricultural
  "Background"                = 0,   # Non-agricultural (assumed background noise)
  "Herbaceous Wetlands"       = 0,   # Non-agricultural
  "Other Hay/Non Alfalfa"     = 1,   # Agricultural
  "Grassland/Pasture"         = 1,   # Agricultural
  "Open Water"                = 0,   # Non-agricultural
  "Barren"                    = 0,   # Non-agricultural
  "Soybeans"                  = 1,   # Agricultural
  "Apples"                    = 1,   # Agricultural
  "Evergreen Forest"          = 0,   # Non-agricultural
  "Shrubland"                 = 0,   # Non-agricultural
  "Dbl Crop WinWht/Corn"      = 1,   # Agricultural
  "Fallow/Idle Cropland"      = 1,   # Agricultural
  "Alfalfa"                   = 1,   # Agricultural
  "Dbl Crop Triticale/Corn"   = 1,   # Agricultural
  "Dbl Crop WinWht/Soybeans"  = 1,  # Agricultural
  "Blueberries"               = 1,  # Agricultural
  "Oats"                      = 1,  # Agricultural
  "Tomatoes"                  = 1,  # Agricultural
  "Christmas Trees"           = 1,  # Agricultural (assuming a crop)
  "Sod/Grass Seed"            = 1,  # Agricultural
  "Winter Wheat"              = 1,  # Agricultural
  "Sorghum"                   = 1,  # Agricultural
  "Dbl Crop Barley/Corn"      = 1,  # Agricultural
  "Canola"                    = 1,  # Agricultural
  "Rye"                       = 1   # Agricultural
)

# Apply the numeric mapping to the dataset
gbif_data <- gbif_data %>%
  mutate(
    nlcd_numeric = recode(nlcd_value, !!!nlcd_land_use_codes, .default = NA_real_),  # Use NA for unmatched values
    nass_numeric = recode(nass_value, !!!nass_land_use_codes, .default = NA_real_)   # Use NA for unmatched values
  )

# Reclassify observations as Agricultural (1) or Non-Agricultural (0)
gbif_data <- gbif_data %>%
  mutate(
    ag_classification = case_when(
      nlcd_numeric == 1 ~ "Agricultural",          # NLCD Agricultural classification
      nass_numeric %in% 1 ~ "Agricultural",     # NASS Agricultural classifications (assuming 1:19 are crops)
      TRUE ~ "Non-Agricultural"                    # Everything else
    )
  )

# Classify agreement/disagreement between NLCD and NASS
gbif_data <- gbif_data %>%
  mutate(
    ag_disagreement = case_when(
      nlcd_numeric == 1 & nass_numeric == 1 ~ "Agree - Agricultural",  # Both agree it's agricultural
      nlcd_numeric == 0 & nass_numeric == 0 ~ "Agree - Non-Agricultural",  # Both agree it's non-agricultural
      nlcd_numeric == 1 & nass_numeric == 0 ~ "Disagree - NLCD Agricultural, NASS Non-Agricultural",  # NLCD says agricultural, NASS says non-agricultural
      nlcd_numeric == 0 & nass_numeric == 1 ~ "Disagree - NLCD Non-Agricultural, NASS Agricultural",  # NLCD says non-agricultural, NASS says agricultural
      TRUE ~ "Unclassified"  # Handle any remaining cases
    )
  )
```
## Look at where the datasets agree and disagree
```{r}
# Calculate the proportions of agreement/disagreement observations
ag_disagreement_proportions <- gbif_data %>%
  group_by(ag_disagreement) %>%
  summarise(count = n()) %>%
  mutate(proportion = count / sum(count))

# Print proportions to verify
print(ag_disagreement_proportions)
# Create a pie chart for agreement/disagreement between NLCD and NASS
ggplot(ag_disagreement_proportions, aes(x = "", y = proportion, fill = ag_disagreement)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  labs(title = "Agreement/Disagreement between NLCD and NASS on Land Use Classification") +
  theme_void() +
  scale_fill_manual(values = c(
    "Agree - Agricultural" = "#66c2a5", 
    "Agree - Non-Agricultural" = "#fc8d62",
    "Disagree - NLCD Agricultural, NASS Non-Agricultural" = "#8da0cb", 
    "Disagree - NLCD Non-Agricultural, NASS Agricultural" = "#e78ac3",
    "Unclassified" = "#a6d854"  # Optional color for unclassified cases
  ))
```


## Calculate proportions of ag vs non ag
```{r}
# Calculate the proportions of agricultural vs non-agricultural observations
ag_proportions <- gbif_data %>%
  group_by(ag_classification) %>%
  summarise(count = n()) %>%
  mutate(proportion = count / sum(count))

# Print proportions to verify
print(ag_proportions)
```


# Visualizations

## Agreed ag vs non-ag

```{r}
# Filter for observations where NLCD and NASS agree on agricultural vs non-agricultural
agreed_data <- gbif_data %>%
  filter(
    (nlcd_numeric == 1 & nass_numeric == 1) |  # Both agree it's agricultural
    (nlcd_numeric == 0 & nass_numeric == 0)    # Both agree it's non-agricultural
  ) %>%
  mutate(
    agreed_classification = ifelse(nlcd_numeric == 1, "Agricultural", "Non-Agricultural")
  )

# Calculate the proportions of agreed upon agricultural vs non-agricultural observations
agreed_proportions <- agreed_data %>%
  group_by(agreed_classification) %>%
  summarise(count = n()) %>%
  mutate(
    proportion = count / sum(count),
    percentage_label = paste0(round(proportion * 100, 1), "%")  # Create percentage labels
  )

# Print proportions to verify
print(agreed_proportions)

# Create a pie chart for agreed agricultural vs non-agricultural observations with labels
ggplot(agreed_proportions, aes(x = "", y = proportion, fill = agreed_classification)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  geom_text(aes(label = percentage_label), 
            position = position_stack(vjust = 0.5),  # Position labels in the middle of each slice
            color = "white", size = 5) +  # Adjust text size and color as needed
  labs(title = "Proportion of Agreed Agricultural vs Non-Agricultural Observations") +
  theme_void() +
  scale_fill_manual(values = c("Agricultural" = "#66c2a5", "Non-Agricultural" = "#fc8d62")) # Optional color scheme
```

## Agricultural observations by crop type
```{r}
# Filter for observations classified as agricultural based on NASS
agricultural_observations <- gbif_data %>%
  filter(nass_numeric > 0)  # Keep only agricultural observations

# Count the number of observations for each NASS value
nass_counts <- agricultural_observations %>%
  group_by(nass_value) %>%
  summarise(count = n()) %>%
  arrange(desc(count))  # Sort by count

# Print counts to verify
print(nass_counts)

# Create a bar plot for the number of agricultural observations by NASS value
ggplot(nass_counts, aes(x = reorder(nass_value, count), y = count, fill = as.factor(nass_value))) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Agricultural Observations by NASS Value",
    x = "NASS Value",
    y = "Number of Observations"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  scale_fill_discrete(name = "NASS Value")  # Include legend for NASS values
```







