---
title: "R Notebook"
output: html_notebook
---

Load Dataset

```{r}
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")

data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
```

# The percentage of observations made on agricultural lands

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)
library(leaflet)
library(leaflet.extras)

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Convert FIPS in data_filtered to character to match the GeoJSON
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Perform the join
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Plot the choropleth map
ggplot(ny_counties_data) +
  geom_sf(aes(fill = Sum_Percent_GBIF_Observations_Agricultural), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Percent on Ag Land") +
  theme_minimal() +
  labs(title = "GBIF Observations on Agricultural Land (2023)",
       caption = "Figure X. Percentage of GBIF occurrence observations taken on agricultural lands\nby county in New York State. Data sources: Global Biodiversity Information Facility (GBIF, 2024),\nNational Land Cover Database (NLCD, 2023), and National Weather Service (NWS, 2024).",
       fill = "Agricultural %") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.caption = element_text(size = 8, hjust = 0),  # Adjust font size and alignment of the caption
    legend.position = "right",
    plot.margin = margin(10, 10, 10, 10)  # Add margins around the plot
  )

# Interactive Map with Leaflet
# Transform to WGS84 (needed for Leaflet)
ny_counties_data <- st_transform(ny_counties_data, crs = 4326)

# Create a color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = ny_counties_data$Sum_Percent_GBIF_Observations_Agricultural,
  na.color = "transparent"
)

# Create the interactive map
leaflet(ny_counties_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(Sum_Percent_GBIF_Observations_Agricultural),
    weight = 1,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste(
      "County:", COUNTYNAME.x, "-",
      "Percent Ag Land Observations:", Sum_Percent_GBIF_Observations_Agricultural, "%"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~Sum_Percent_GBIF_Observations_Agricultural,
    title = "Percent on Ag Land",
    opacity = 1
  ) %>%
  addFullscreenControl()
```

## Bar Plot

```{r}
# Bar plot assigned to a variable
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, -Sum_Percent_GBIF_Observations_Agricultural), 
                                      y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +  # Add margin between bars
  coord_flip() +
  theme_minimal() +
  labs(title = "GBIF Observations in Agricultural Land by NY County (2023)",
       x = "County", y = "Percentage of Observations",
       caption = "") +
  theme(
    axis.text.y = element_text(size = 10)  # Increase text size for readability
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_with_margins.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Boxplot

```{r}
ggplot(data_filtered, aes(y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(
    title = "Distribution of GBIF Agricultural Observations by County (2023)",
    y = "Percentage of Observations on Agricultural Lands",
    caption = ""
  ) +
  theme(
    axis.title.x = element_blank(),  # Remove x-axis label
    axis.text.x = element_blank(),   # Remove x-axis tick labels
    axis.ticks.x = element_blank()   # Remove x-axis ticks
  )
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_histogram(binwidth = .1, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of GBIF Observations on Agricultural Land (2023)",
       x = "Percentage of Observations",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```

## Scatterplot

```{r}
# Scatter plot
ggplot(data_filtered, aes(x = X2023_GBIF_Observation_Count, y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_point(color = "purple", size = 3, alpha = 0.7) +
  theme_minimal() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Scatter Plot: Agricultural Observation Percentage vs Total Observations",
       x = "Total GBIF Observations by County",
       y = "Percentage of Observations in Agricultural Land",
       caption = "Data Source: GBIF 2023")
```

# Observational Disparity
(Total Percent GBIF Observations on Agricultural Lands / Total Percent of Agricultural Land)

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)
library(leaflet)
library(leaflet.extras)
library(ggrepel)  # For better label placement

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Read and filter the CSV file
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Join spatial and observational data
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Filter counties where GBIF_Agricultural_Disparity is greater than 1
labeled_counties <- ny_counties_data %>%
  filter(GBIF_Agricultural_Disparity > 1) %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(x = st_coordinates(centroid)[,1], 
         y = st_coordinates(centroid)[,2])

# ---- STATIC MAP ----
ggplot(ny_counties_data) +
  geom_sf(aes(fill = GBIF_Agricultural_Disparity), color = "black", size = 0.2) +

  # Label only counties with greater than 1 disparity
  geom_label_repel(data = labeled_counties, 
                   aes(x = x, y = y, label = COUNTYNAME.x), 
                   size = 4, fontface = "bold", 
                   fill = "white", color = "black",
                   box.padding = 0.5, point.padding = 0.3, segment.color = "white") + 
  
  scale_fill_viridis_c(option = "plasma", name = "") +
  theme_minimal() +
  labs(title = "",
       caption = "",
       fill = "",
       x = NULL,  # Removes x-axis label
       y = NULL) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right",
    axis.text = element_blank(),       # Remove axis text
    axis.ticks = element_blank(),      # Remove axis ticks
    panel.grid = element_blank()       # Remove background grid lines
  )

# ---- INTERACTIVE MAP ----
# Transform to WGS84 (needed for Leaflet)
ny_counties_data <- st_transform(ny_counties_data, crs = 4326)
labeled_counties <- st_transform(labeled_counties, crs = 4326)

# Create a color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = ny_counties_data$GBIF_Agricultural_Disparity,
  na.color = "transparent"
)

# Create the interactive map with labeled counties
leaflet(ny_counties_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(GBIF_Agricultural_Disparity),
    weight = 1,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste(
      "County:", COUNTYNAME.x, "<br>",
      "Observational Disparity:", round(GBIF_Agricultural_Disparity, 2)
    ),
    labelOptions = labelOptions(
      style = list("font-size" = "12px", "background" = "white", "padding" = "5px")
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~GBIF_Agricultural_Disparity,
    title = "Observational Disparity",
    opacity = 1
  ) %>%
  addFullscreenControl() %>%
  addLabelOnlyMarkers(
    data = labeled_counties,
    lng = ~x, lat = ~y,
    label = ~COUNTYNAME.x,
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "top",
      textOnly = TRUE,
      style = list(
        "color" = "black",
        "font-weight" = "bold",
        "background" = "white",
        "border" = "2px solid black",
        "padding" = "5px",
        "border-radius" = "5px"
      )
    )
  )
```

## Bar Plot

```{r}
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, -GBIF_Agricultural_Disparity), 
                                      y = GBIF_Agricultural_Disparity, 
                                      fill = GBIF_Agricultural_Disparity < 1)) + 
  geom_bar(stat = "identity", width = 0.7) +  
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "steelblue")) +  # Assign colors
  coord_flip() +
  theme_minimal() +
  labs(title = "",
       x = "", y = "",
       caption = "") +
  theme(
    axis.text.y = element_text(size = 10),  # Increase text size for readability
    legend.position = "none",  # Remove legend if not needed
    panel.grid.major = element_blank(),     # Remove major gridlines
    panel.grid.minor = element_blank() 
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_with_margins.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Box Plot

```{r}
# Box plot
ggplot(data_filtered, aes(y = GBIF_Agricultural_Disparity)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "Distribution of GBIF Agricultural Disparity",
       y = "% Ag Obs / % Ag Total",
       caption = "Data Source: GBIF 2023")
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = GBIF_Agricultural_Disparity)) +
  geom_histogram(binwidth = 0.25, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of GBIF Agricultural Disparity",
       x = "% Ag Obs / % Ag Total",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```

# Observational Coverage

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)
library(leaflet)
library(leaflet.extras)
library(ggrepel)  # For automatic non-overlapping labels

# Read the GeoJSON file (New York counties)
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Read and filter the CSV file (Observational data)
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Join spatial and observational data
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Calculate the complement of Percent_Agricultural_Land_Greater_Than_1km_GBIF
ny_counties_data <- ny_counties_data %>%
  mutate(Observational_Coverage = 1 - Percent_Agricultural_Land_Greater_Than_1km_GBIF)

# Identify the top 5 highest and bottom 5 lowest counties
top_5_counties <- ny_counties_data %>%
  arrange(desc(Observational_Coverage)) %>%
  head(5)

bottom_5_counties <- ny_counties_data %>%
  arrange(Observational_Coverage) %>%
  head(5)

# Merge top and bottom 5 counties
highlighted_counties <- bind_rows(top_5_counties, bottom_5_counties)

# Compute centroids for labeling
highlighted_counties <- highlighted_counties %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(x = st_coordinates(centroid)[,1], 
         y = st_coordinates(centroid)[,2])

# ---- STATIC MAP ----
ggplot(ny_counties_data) +
  geom_sf(aes(fill = Observational_Coverage), color = "black", size = 0.2) +

  # Use only geom_label_repel() to avoid double leader lines
  geom_label_repel(data = highlighted_counties, 
                   aes(x = x, y = y, label = COUNTYNAME.x), 
                   size = 4, fontface = "bold", 
                   fill = "white", color = "black",  # High contrast for all backgrounds
                   box.padding = 0.5, point.padding = 0.3, segment.color = "white") + 
  
  scale_fill_viridis_c(option = "plasma", name = "") +
  theme_minimal() +
  labs(title = "",
       caption = "",
       fill = "Agricultural %",
       x = NULL,  # Removes x-axis label
       y = NULL) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right",
    axis.text = element_blank(),       # Remove axis text
    axis.ticks = element_blank(),      # Remove axis ticks
    panel.grid = element_blank()       # Remove background grid lines
  )

# ---- INTERACTIVE MAP ----
# Create a color palette
pal <- colorNumeric(
  palette = "plasma",
  domain = ny_counties_data$Observational_Coverage
)

leaflet(ny_counties_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(Observational_Coverage),
    weight = 1,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.9,
      bringToFront = TRUE
    ),
    label = ~paste(
      "County:", COUNTYNAME.x, "<br>",
      "Observational Coverage:", round(Observational_Coverage, 3)
    ),
    labelOptions = labelOptions(
      style = list("font-size" = "12px", "background" = "white", "padding" = "5px")
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~Observational_Coverage,
    title = "Observational Coverage",
    opacity = 1
  ) %>%
  addFullscreenControl() %>%
  addLabelOnlyMarkers(
    data = highlighted_counties,
    lng = ~x, lat = ~y,
    label = ~COUNTYNAME.x,
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "top",
      textOnly = TRUE,
      style = list(
        "color" = "black",  # Ensures visibility on all backgrounds
        "font-weight" = "bold",
        "background" = "white",
        "border" = "2px solid black",  # Ensures high contrast
        "padding" = "5px",
        "border-radius" = "5px"
      )
    )
  )
```

## Bar Plot

```{r}
library(sf)
library(dplyr)
library(ggplot2)

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Read and filter the CSV file
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Calculate the complement of Percent_Agricultural_Land_Greater_Than_1km_GBIF
data_filtered <- data_filtered %>%
  mutate(Observational_Coverage = 1 - Percent_Agricultural_Land_Greater_Than_1km_GBIF)

# Identify the top 5 highest and bottom 5 lowest counties
top_5_counties <- data_filtered %>%
  arrange(desc(Observational_Coverage)) %>%
  head(5) %>%
  pull(COUNTYNAME)

bottom_5_counties <- data_filtered %>%
  arrange(Observational_Coverage) %>%
  head(5) %>%
  pull(COUNTYNAME)

# Add a new column to classify counties into top 5, bottom 5, or others
data_filtered <- data_filtered %>%
  mutate(Highlight = case_when(
    COUNTYNAME %in% top_5_counties ~ "Top 5",
    COUNTYNAME %in% bottom_5_counties ~ "Bottom 5",
    TRUE ~ "Others"
  ))

# Bar plot with highlighted top 5 and bottom 5 counties
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, Observational_Coverage), 
                                      y = Observational_Coverage, 
                                      fill = Highlight)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = c("Top 5" = "orange", "Bottom 5" = "purple", "Others" = "gray")) + # Custom colors
  coord_flip() +
  theme_minimal() +
  labs(title = "",
       x = "", y = "",
       caption = "") +
  theme(
    axis.text.y = element_text(size = 10),  # Increase text size for readability
    legend.position = "none",  # Remove legend if not needed
    panel.grid.major = element_blank(),     # Remove major gridlines
    panel.grid.minor = element_blank() 
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_highlighted_top_bottom.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Box Plot

```{r}
# Box plot
ggplot(data_filtered, aes(y = Percent_Agricultural_Land_Near_GBIF)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "Neglected GBIF Coverage",
       y = "Ag Land > 1km GBIF",
       caption = "Data Source: GBIF 2023")
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = Percent_Agricultural_Land_Near_GBIF)) +
  geom_histogram(binwidth = .1, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of Neglected GBIF Coverage",
       x = "Ag Land > 1km GBIF",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```
# General Statistics

## Observational Density
```{r}
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)

# Calculate the median of the specified column in the filtered dataset
median_value <- median(data_filtered$X2023_GBIF_Observation_Count_Per_km2, na.rm = TRUE)

print(paste('Median observations per km2 in 2023: ', median_value))
print(paste('Median observations per km2 per month in 2023: ', median_value / 12))
print(paste('Median obs/100km²/month in 2023: ', (median_value / 12) * 100))
```
```{r}
install.packages("psych")
# Load required libraries
library(ggplot2)
library(dplyr)
library(psych)  # For descriptive statistics

# Read the dataset
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")

# Filter the data
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)

# Assign variables for analysis
population_density <- data_filtered$X2023.Population.Density.km2
proportion_gbif_observations_ag <- data_filtered$GBIF_Agricultural_Disparity
proportion_ag_lands_not_near_gbif_observations <- data_filtered$Percent_Agricultural_Land_Near_GBIF

# ---------------------------------
# 4. Scatterplots with Trendlines
# ---------------------------------
cat("\nGenerating Scatterplots...\n")

# Scatterplot 1: median_household_income vs GBIF Disparity
ggplot(data_filtered, aes(x = population_density, y = proportion_gbif_observations_ag)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Population Density vs GBIF Agricultural Disparity",
       x = "Population Density km2",
       y = "GBIF Agricultural Disparity")

# Scatterplot 2: Unemployment Levels vs Percent Ag Lands Near GBIF
ggplot(data_filtered, aes(x = population_density, y = proportion_ag_lands_not_near_gbif_observations)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "green") +
  labs(title = "Population Density vs Percent Ag Lands >1km GBIF",
       x = "Population Density km2",
       y = "Percent Agricultural Land >1km GBIF")

# ---------------------------------
# 5. Regression Analysis (Optional)
# ---------------------------------
cat("\nRunning Regression Models...\n")

# Regression 1: Education Levels -> GBIF Agricultural Disparity
model1 <- lm(proportion_gbif_observations_ag ~ population_density, data = data_filtered)
summary_model1 <- summary(model1)
print("Regression - Population Density predicting GBIF Disparity")
print(summary_model1)

# Regression 2: Education Levels -> Percent Ag Lands Near GBIF
model2 <- lm(proportion_ag_lands_not_near_gbif_observations ~ population_density, data = data_filtered)
summary_model2 <- summary(model2)
print("Regression - Population Density predicting Percent Ag Lands >1km GBIF")
print(summary_model2)
```

