---
title: "R Notebook"
output: html_notebook
---

Load Dataset

```{r}
data <- read.csv("../Dataset/v2_2023_gbif_ny.csv")

data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
```

# The percentage of observations made on agricultural lands

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Convert FIPS in data_filtered to character to match the GeoJSON
data <- read.csv("../Dataset/v1_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Perform the join
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Plot the choropleth map
ggplot(ny_counties_data) +
  geom_sf(aes(fill = Sum_Percent_GBIF_Observations_Agricultural), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "Percent in Ag Land") +
  theme_minimal() +
  labs(title = "GBIF Observations in Agricultural Land (2023)",
       caption = "Data Source: GBIF 2023",
       fill = "Agricultural %") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

## Bar Plot

```{r}
# Bar plot assigned to a variable
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, -Sum_Percent_GBIF_Observations_Agricultural), 
                                      y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +  # Add margin between bars
  coord_flip() +
  theme_minimal() +
  labs(title = "GBIF Observations in Agricultural Land by NY County (2023)",
       x = "County", y = "Percentage of Observations",
       caption = "Data Source: GBIF 2023") +
  theme(
    axis.text.y = element_text(size = 10)  # Increase text size for readability
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_with_margins.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Boxplot

```{r}
# Box plot
ggplot(data_filtered, aes(y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "Distribution of GBIF Agricultural Observations by County (2023)",
       y = "Percentage of Observations",
       caption = "Data Source: GBIF 2023")
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_histogram(binwidth = .1, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of GBIF Observations in Agricultural Land (2023)",
       x = "Percentage of Observations",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```

## Scatterplot

```{r}
# Scatter plot
ggplot(data_filtered, aes(x = X2023_GBIF_Observation_Count, y = Sum_Percent_GBIF_Observations_Agricultural)) +
  geom_point(color = "purple", size = 3, alpha = 0.7) +
  theme_minimal() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Scatter Plot: Agricultural Observation Percentage vs Total Observations",
       x = "Total GBIF Observations in County",
       y = "Percentage of Observations in Agricultural Land",
       caption = "Data Source: GBIF 2023")
```

# GBIF Agricultural Disparity (Total Percent GBIF Observations on Agricultural Lands / Total Percent of Agricultural Land)

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Convert FIPS in data_filtered to character to match the GeoJSON
data <- read.csv("../Dataset/v1_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Perform the join
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Plot the choropleth map
ggplot(ny_counties_data) +
  geom_sf(aes(fill = GBIF_Agricultural_Disparity), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "% Ag Obs / % Ag Total") +
  theme_minimal() +
  labs(title = "GBIF Agricultural Disparity",
       caption = "Data Source: GBIF 2023",
       fill = "Agricultural %") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

## Bar Plot

```{r}
# Bar plot assigned to a variable
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, -GBIF_Agricultural_Disparity), 
                                      y = GBIF_Agricultural_Disparity)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +  # Add margin between bars
  coord_flip() +
  theme_minimal() +
  labs(title = "GBIF Agricultural Disparity",
       x = "County", y = "% Ag Obs / % Ag Total",
       caption = "Data Source: GBIF 2023, NLCD") +
  theme(
    axis.text.y = element_text(size = 10)  # Increase text size for readability
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_with_margins.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Box Plot

```{r}
# Box plot
ggplot(data_filtered, aes(y = GBIF_Agricultural_Disparity)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "Distribution of GBIF Agricultural Disparity",
       y = "% Ag Obs / % Ag Total",
       caption = "Data Source: GBIF 2023")
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = GBIF_Agricultural_Disparity)) +
  geom_histogram(binwidth = 0.25, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of GBIF Agricultural Disparity",
       x = "% Ag Obs / % Ag Total",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```

# Percentage of Agricultural Lands not within 1km of GBIF Observations

## Choropleth Map

```{r}
# Load required libraries
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)

# Read the GeoJSON file
ny_counties_geo <- st_read("../Dataset/ny_counties_statistics.geojson")

# Convert FIPS in data_filtered to character to match the GeoJSON
data <- read.csv("../Dataset/v1_2023_gbif_ny.csv")
data_filtered <- subset(data, Sum_Percent_Land_Use_Agricultural > 0.05)
data_filtered$FIPS <- as.character(data_filtered$FIPS)

# Perform the join
ny_counties_data <- ny_counties_geo %>%
  left_join(data_filtered, by = "FIPS")

# Plot the choropleth map
ggplot(ny_counties_data) +
  geom_sf(aes(fill = Percent_Agricultural_Land_Near_GBIF), color = "black", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", name = "% Ag > 1km from Obs.") +
  theme_minimal() +
  labs(title = "GBIF Coverage",
       caption = "Data Source: GBIF 2023",
       fill = "Agricultural %") +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  )
```

## Bar Plot

```{r}
# Bar plot assigned to a variable
bar_plot <- ggplot(data_filtered, aes(x = reorder(COUNTYNAME, -Percent_Agricultural_Land_Near_GBIF), 
                                      y = Percent_Agricultural_Land_Near_GBIF)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) +  # Add margin between bars
  coord_flip() +
  theme_minimal() +
  labs(title = "Neglected GBIF Coverage",
       x = "County", y = "Ag Land > 1km GBIF",
       caption = "Data Source: GBIF 2023, NLCD") +
  theme(
    axis.text.y = element_text(size = 10)  # Increase text size for readability
  )

# Display the plot
print(bar_plot)

# Save the plot
ggsave("bar_plot_with_margins.png", plot = bar_plot, width = 8, height = 8, dpi = 300)
```

## Box Plot

```{r}
# Box plot
ggplot(data_filtered, aes(y = Percent_Agricultural_Land_Near_GBIF)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "Neglected GBIF Coverage",
       y = "Ag Land > 1km GBIF",
       caption = "Data Source: GBIF 2023")
```

## Histogram

```{r}
# Histogram
ggplot(data_filtered, aes(x = Percent_Agricultural_Land_Near_GBIF)) +
  geom_histogram(binwidth = .1, fill = "orange", color = "black") +  # Smaller binwidth
  theme_minimal() +
  labs(title = "Histogram of Neglected GBIF Coverage",
       x = "Ag Land > 1km GBIF",
       y = "Count of Counties",
       caption = "Data Source: GBIF 2023")
```
